---
title: "Getting Started with Trex"
author:
- name: Nick Borcherding
  email: ncborch@gmail.com
  affiliation: Washington University in St. Louis, School of Medicine, St. Louis, MO, USA
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc_float: true
package: Trex
vignette: >
  %\VignetteEngine{knitr::knitr}
  %\VignetteIndexEntry{Getting Started with Trex}
  %\usepackage[UTF-8]{inputenc}
---

```{r, echo=FALSE, results="hide", message=FALSE}
knitr::opts_chunk$set(error = FALSE, message = FALSE, warning = FALSE)
library(BiocStyle)
```

# Introduction

Single-cell sequencing is now an integral tool in the field of immunology and
oncology that allows researchers to couple RNA quantification and other
modalities, like immune cell receptor profiling at the level of an individual
cell. Towards this end, we developed the
[scRepertoire](https://github.com/BorchLab/scRepertoire) R package to assist in
the interaction of immune receptor and gene expression sequencing.
However, utilization of clonal indices for more complex analyses are still
lacking, specifically in using clonality in embedding of single-cells. To this
end, we developed Trex - an R package that uses deep learning to vectorize TCR
sequences using sequence order or translating the sequence into amino acid
properties.

## Installation

Trex is available through Bioconductor:

```{r, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}
BiocManager::install("Trex")
```

## Model Downloads

By default, Trex will automatically download deep learning models from a
[Zenodo repository](https://zenodo.org/records/14636032) and cache them locally
on first use. The cache location follows platform conventions using
`tools::R_user_dir("Trex", which = "cache")`.

## Load Libraries

```{r}
suppressMessages(library(Seurat))
suppressMessages(library(ggplot2))
suppressMessages(library(Trex))
suppressMessages(library(viridis))
suppressMessages(library(patchwork))
```

# The Data Set

To demonstrate Trex's capabilities, the example data is derived from
[GSE167118](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE167118) - a
cohort of CITE-seq data from severe COVID-19 patients. The
[corresponding manuscript](https://pubmed.ncbi.nlm.nih.gov/33622974/) is
excellent. The built-in data (`trex_example`) is a random sample of single
cells from patient 17.

This vignette follows a standard workflow based on the
[WNN](https://satijalab.org/seurat/articles/weighted_nearest_neighbor_analysis.html)
Seurat process. Trex also works with Bioconductor/SingleCellExperiment
workflows.

## Loading the Data

```{r}
data("trex_example")
```

# Running Trex

The idea behind Trex is to combine TCR CDR3 amino acid information with
phenotypic RNA/protein data to direct single-cell sequencing towards
antigen-specific discoveries. This is a growing field - specifically
[TESSA](https://github.com/jcao89757/TESSA) uses amino acid characteristics and
autoencoders as a means to get a dimensional reduction. Another option is
[CoNGA](https://github.com/phbradley/conga), which produces an embedding using
TCR and RNA. Trex was designed to make a customizable approach to this combined
embedding using R.

## The maTrex Function

Trex has two core functions - the first being `maTrex()`, which is the
backbone of the algorithm and returns the encoded values based on the selection
of variables. Unlike `runTrex()` below, `maTrex()` does not filter the input
for only T cells with attached TCR data. In addition, `maTrex()` is compatible
with the list output from the `combineTCR()` function from the
[scRepertoire](https://github.com/BorchLab/scRepertoire) R package, while
`runTrex()` must be performed on a single-cell object.

### Parameters

**chains**
TCR chain to use for encoding:

* `"TRA"`: T cell receptor alpha chain
* `"TRB"`: T cell receptor beta chain

**method**
Encoding method to use:

* `"encoder"`: Deep learning autoencoders for sequence embedding
* `"geometric"`: Geometric transformation based on BLOSUM62 substitution matrix

**encoder.model**
Model architecture when `method = "encoder"`:

* `"VAE"`: Variational autoencoder
* `"AE"`: Dense autoencoder

**encoder.input**
Input encoding when `method = "encoder"`:

* `"AF"`: Atchley factors (5 physicochemical properties)
* `"KF"`: Kidera factors (10 physicochemical properties)
* `"OHE"`: One-hot encoding (21 dimensions per position)

**theta**
Rotation angle in radians for geometric transformation when
`method = "geometric"`. Default is `pi`.

### Examples

```{r}
Trex_vectors <- maTrex(trex_example,
    chains = "TRA",
    method = "encoder",
    encoder.model = "VAE",
    encoder.input = "AF"
)

ggplot(data = as.data.frame(Trex_vectors), aes(Trex_1, Trex_2)) +
    geom_point() +
    theme_classic()

Trex_vectors2 <- maTrex(trex_example,
    chains = "TRB",
    method = "geometric",
    theta = pi
)

ggplot(as.data.frame(Trex_vectors2), aes(x = Trex_1, y = Trex_2)) +
    geom_point() +
    theme_classic()
```

## The runTrex Function

The `runTrex()` function appends the Seurat or SingleCellExperiment object with
the Trex vectors and allows for further analysis. Importantly, `runTrex()` will
remove single cells that do not have recovered TCR data in the metadata of the
object.

```{r}
trex_example <- runTrex(trex_example,
    chains = "TRB",
    method = "encoder",
    encoder.model = "VAE",
    encoder.input = "KF",
    reduction.name = "Trex.KF"
)
```

## Using Trex Vectors

After `runTrex()` we have the encoded values stored under **"Trex..."**. Using
the Trex reduction stored in Seurat, we can calculate the nearest neighbor and
shared nearest neighbor indexes and generate a UMAP.

```{r}
# Generating UMAP from Trex Neighbors
trex_example <- RunUMAP(trex_example,
    reduction = "Trex.KF",
    dims = 1:30,
    reduction.name = "Trex.umap",
    reduction.key = "trexUMAP_"
)

# Trex UMAP
plot1 <- DimPlot(trex_example, reduction = "Trex.umap") + NoLegend()
plot2 <- DimPlot(trex_example, group.by = "CTaa", reduction = "Trex.umap") +
    scale_color_viridis(discrete = TRUE, option = "B") +
    NoLegend()

plot1 + plot2
```

We can now use this in a similar way as other single-cell modalities and
calculate weighted nearest neighbor (WNN). To learn more about WNN, please read
the Satija group's [paper](https://pubmed.ncbi.nlm.nih.gov/34062119/). We will
use the RNA, ADT protein levels, and Trex vectors for the WNN calculations.

```{r}
trex_example <- FindMultiModalNeighbors(
    trex_example,
    reduction.list = list("pca", "apca", "Trex.KF"),
    dims.list = list(1:30, 1:20, 1:30),
    modality.weight.name = "RNA.weight"
)

trex_example <- RunUMAP(trex_example,
    nn.name = "weighted.nn",
    reduction.name = "wnn.umap",
    reduction.key = "wnnUMAP_"
)

trex_example <- FindClusters(trex_example,
    graph.name = "wsnn",
    resolution = 0.6,
    algorithm = 3,
    verbose = FALSE
)

# WNN UMAP
plot3 <- DimPlot(trex_example, reduction = "wnn.umap")
plot4 <- DimPlot(trex_example, reduction = "wnn.umap", group.by = "CTaa") +
    scale_color_manual(values = viridis_pal(option = "B")(
        length(unique(trex_example$CTaa))
    )) +
    NoLegend()

plot3 + plot4
```

## Comparing Modalities

We can also look at the differences in the UMAP generated from RNA, ADT, or
Trex as individual components. Remember, the clusters displayed in UMAP are
based on clusters defined by the weighted nearest neighbors calculated above.

```{r}
trex_example <- RunUMAP(trex_example,
    reduction = "pca",
    dims = 1:30,
    assay = "RNA",
    reduction.name = "rna.umap",
    reduction.key = "rnaUMAP_"
)

trex_example <- RunUMAP(trex_example,
    reduction = "apca",
    dims = 1:20,
    assay = "ADT",
    reduction.name = "adt.umap",
    reduction.key = "adtUMAP_"
)

plot5 <- DimPlot(trex_example, reduction = "rna.umap") + NoLegend()
plot6 <- DimPlot(trex_example, reduction = "adt.umap") + NoLegend()
plot7 <- DimPlot(trex_example, reduction = "Trex.umap") + NoLegend()

plot5 + plot6 + plot7
```

# CoNGA Reduction

Recent [work](https://pubmed.ncbi.nlm.nih.gov/34426704/) has proposed using
representative cells for the characterization of clonotype and gene expression
relationships. In order to generate these representative cells, either a mean
expression across a clone or using the PCA dimensional space to identify a
single cell that has the minimum Euclidean distance across a clone.

In order to generate a single-cell object based on the CoNGA approach, Trex
offers the function `CoNGAfy()`. For **method**, select either `"mean"` or
`"dist"` as described above. After performing `CoNGAfy()`, the user can use any
of the above reduction strategies.

```{r}
library(dplyr, include.only = c("%>%"))
CoNGA.seurat <- CoNGAfy(trex_example,
    method = "dist"
)

CoNGA.seurat <- runTrex(CoNGA.seurat,
    chains = "TRB",
    method = "encoder",
    encoder.model = "VAE",
    encoder.input = "KF",
    reduction.name = "Trex.KF"
)

CoNGA.seurat <- CoNGA.seurat %>%
    FindNeighbors(reduction = "Trex.KF") %>%
    FindClusters(algorithm = 3, resolution = 0.9)

CoNGA.seurat <- RunUMAP(CoNGA.seurat,
    reduction = "Trex.KF",
    dims = 1:20,
    reduction.name = "Trex.umap",
    reduction.key = "trexUMAP_"
)

DimPlot(CoNGA.seurat, reduction = "Trex.umap") + NoLegend()
```

# Annotating Epitopes

To find epitope relationships, Trex has a built-in database of TRA and TRB
sequences associated with epitopes. To append the database to the single-cell
object (either before or after `CoNGAfy()`), use `annotateDB()`.

The database is collated from multiple sources:

1. [VDJdb](https://vdjdb.cdr3.net/) - [citation](https://pubmed.ncbi.nlm.nih.gov/28977646/)
2. [McPAS-TCR](http://friedmanlab.weizmann.ac.il/McPAS-TCR/) - [citation](https://pubmed.ncbi.nlm.nih.gov/28481982/)
3. [IEDB](https://www.iedb.org/) - [citation](https://pubmed.ncbi.nlm.nih.gov/30357391/)
4. [PIRD](https://db.cngb.org/pird/home/) - [citation](https://pubmed.ncbi.nlm.nih.gov/31373607/)

```{r}
CoNGA.seurat <- annotateDB(CoNGA.seurat,
    chains = "TRB"
)

DimPlot(CoNGA.seurat,
    reduction = "Trex.umap",
    group.by = "TRB_Epitope.species"
)

DimPlot(CoNGA.seurat,
    reduction = "Trex.umap",
    group.by = "TRB_Epitope.sequence"
)
```

Or `annotateDB()` can be used on the full single-cell object and examine the
sequence information along the RNA-based UMAP. An added feature allows the
annotations to be extended to CDR3 sequences within defined edit distances from
the reference using **edit.distance**.

```{r}
trex_example <- annotateDB(trex_example,
    chains = "TRB",
    edit.distance = 2
)

DimPlot(trex_example,
    reduction = "wnn.umap",
    group.by = "TRB_Epitope.species"
)
```

# Removing TCR Genes

When combining TCR-based embeddings with RNA-based embeddings using weighted
nearest neighbors (WNN), it is important to remove TCR genes from the variable
features before running PCA. The `quietTCRgenes()` function removes T cell
receptor genes (matching pattern `^TR[ABDG][VDJ][^D]`) from the variable
features in a Seurat object.
```{r, eval=FALSE}
# Remove TCR genes before running PCA for WNN analysis
seurat_obj <- quietTCRgenes(seurat_obj)
seurat_obj <- RunPCA(seurat_obj)
```

# Conclusion

This has been a general overview of the capabilities of Trex for incorporating
TCR information into the embedding space of single-cell data. If you have any
questions, comments, or suggestions, feel free to visit the
[GitHub repository](https://github.com/BorchLab/Trex).

## Session Info

```{r}
sessionInfo()
```
